Consistency and Fault Tolerance Considerations for the Next Iteration of the DOE Fast Forward Storage and IO Project

### 2014-03-22 paper.tex
submission version

\addtolength{\parskip}{-0.04in}

@@

\begin{abstract}
With phase 1 of the Fast Forward Storage and IO Stack project complete, it is
an excellent opportunity to evaluate many of the decisions made to feed into
the phase 2 effort. The initial effort to define a next generation storage
system has made admirable contributions in architecture and design. Formalizing
the general idea of data staging into burst buffers for the storage system will
help manage the performance variability and offer additional data processing
opportunities outside the main compute and storage system. Adding a
transactional mechanism to manage faults and data visibility helps enable
effective analytics without having to work around the IO stack semantics.

While these and other contributions are valuable, similar efforts made
elsewhere may offer attractive alternatives or differing semantics that could
yield a more feature rich environment with little to no additional overhead.
For example, the Doubly Distributed Transactions (\DDTns) protocol offers an
alternative approach for incorporating transactional semantics into the data
path. Another project, PreDatA, examined how to get the best throughput for
data operators and may offer additional insights into further refinements of
the Burst Buffer concept.

This paper examines some of the choices made by the FastForward team and
compares them with other options and offers observations and suggestions based
on these other efforts.  This will include some non-core contributions of other
projects, such as some of the demonstration metadata and data storage
components generated while implementing \DDTns, to make suggestions that may
help the next generation design for how the IO stack works as a whole.

@@

The overall design seeks to offer byte-granular, multi-version concurrency
control. Through the use of a copy-on-write style mechanism, multiple versions
of an object can be stored in potentially greatly reduced space. It assumes the
client interface will be through an IO library affording a more complicated
interface that offers richer functionality requiring only minimal end-user code
changes.  Managing most data access in a platform-local layer rather than
requiring writing to centralized storage will better support the performance
and energy requirements of extreme scale integrated application workflows. At a
more detailed view, the various layers of the IO stack each contribute
different functionality and performance implications.

The basic architecture incorporates five layers. The top layer is a high level
IO library, such as the demonstration HDF-5 library~\cite{hdf5}. The intent is
to only have access to the storage stack through such an API to manage the
complexity of working with the lower layers. Below the user API is an IO
forwarding layer that redirects IO calls from the compute nodes to the IO
dispatching layer.  This IO forwarding layer is analogous to the function of
the IO nodes in a BlueGene machine. The next two layers have considerable
functionality. The IO dispatcher (IOD) serves as the primary storage interface
for the IO stack and is the only way to reach the persistent storage array in
lower layers. The Data Access Object Storage (DAOS) layer serves as the
persistent storage interface and is intended to be the foundation on which
everything else is built with no dependence on any technologies specified above
it. For example, the IOD layer and burst buffers are not required for DAOS to
operate properly. At the bottom is the Versioning Object Storage Device (VOSD).
It serves as the interface for storing objects of all types efficiently.

@@

The idea of burst buffers were initially explored in the context of data
staging~\cite{abbasi:2007:datatap,Abbasi:2009:datatap,nisar:2008:staging,zheng:2010:predata}.
These initial designs all use extra compute nodes to represent the data storage
buffer given the lack of any dedicated hardware support for this functionality.
The desired outcome of these initial studies is to motivate how such
functionality might be incorporated and the potential benefits.  Later, these
concepts were proposed to be incorporated as part of the IO
stack~\cite{nowoczynski:2008:zest,bent:2012:challenges,bent:2012:burst-buffer}.
The current FastForward IOD design recommends incorporating SSDs, but
specifically lists these devices as optional. Unfortunately, not incorporating
burst buffers and the use of SSDs in the IOD layer may be problematic.  First,
the IOD and DAOS designs currently are written assuming burst buffers. This
means that the bulk of IO operations will only hit the IOD layer and proposed
functionality, such as the function shipping, do not discuss the impact on the
DAOS layer or the operators themselves should a burst buffer not be available.
Consider the important functionality of data rearrangement and doing things
like changing the fast array dimension on shared, spinning media.  The design
of DAOS assumes that it will only be involved when persisting a completed
transaction and only for a fraction of the total transactions created.
Transaction frequency at the IOD layer can be higher since IOD does not
``flatten'' transactions. Flattening is the process of transfering the set of
copy-on-write changes from transaction/epoch $n$ to transaction $n+m$ into the
DAOS layer. The DAOS layer mirrors the copy-on-write functionality, but stores
the differences between epochs rather than individual transactions. The primary
difference is that at the DAOS layer, data is stored in large, logically
contiguous chunks to manage the metadata load and hopefully improve read
performance.

@@

By concentrating the Burst Buffers and function shipping into the storage
stack, three problems arise.  First, the amount of network bandwidth, IO
bandwidth, and compute power consumed for example operations from a single
application is likely to completely monopolize the IOD processes. Second, if
space and time partitioning is used instead, the functionality risks being too
small to be useful. Third, the long-term hardware performance advantage for
SSDs is questionable. Recent studies have shown that the erase-before-write
and interference between reading and writing with flash-based SSDs can cause
severe performance problems~\cite{skourtis:2013:ssd-performance}. The inclusion
of an optional use SSD layer in the new Trinity machine at Los Alamos will
offer a testbed to determine how likely these observed problems would affect a
production extreme scale platform.

@@

Another concern that is acknowledged, but no thought has been applied to, is
the requirement that a single IOD process be the master for any operation.
Should the number of concurrent applications exceed the available nodes,
sharing an IOD process will be required. The requirements both in terms of
scheduling and resource management were cut from the project due to funding
limitations. Since partitioning of the IOD processes for exclusive use by
particular applications is the assumed operating mode, should insufficient IOD
resources be available, either a job could be delayed or IOD resources could be
reallocated from a different process could be redeployed for use by the new
job. Handling resilience concerns for the IOD processes must also be address.
These sorts of considerations still need to be made for a full production
system.

@@

\subsubsection{Design Philosophy}
Undocumented, but inherent in the design of these transactions is how faults
are detected. The initial design assumes the current Lustre fault detection
mechanism that can determine if a process or node is no longer reachable. This
defines how a fault will be detected and what will trigger a passive fault
recovery (i.e., transaction abort).

@@

\subsection{Metadata Management}
Metadata management has been a perennial challenge for parallel storage
systems.  Eliminating metadata management as a special case and instead
treating it just as data is a central design goal of the Fast Forward project.
This is a hybrid approach to metadata management that is half-way between
providing no inherent metadata support and having a fully integrated, but
separate metadata manaegement system.

>>>>

\addtolength{\parskip}{-0.06in}

@@

\begin{abstract}
The DOE Extreme-Scale Technology Acceleration Fast Forward Storage and IO Stack
project is going to have significant impact on storage systems design within
and beyond the HPC community. With phase 1 of the project complete, it is an
excellent opportunity to evaluate many of the decisions made to feed into the
phase 2 effort. With this paper we not only provide a timely summary of
important aspects of the design specifications but also capture the underlying
reasoning that is not available elsewhere.

The initial effort to define a next generation storage system has made
admirable contributions in architecture and design.  Formalizing the general
idea of data staging into burst buffers for the storage system will help manage
the performance variability and offer additional data processing opportunities
outside the main compute and storage system. Adding a transactional mechanism
to manage faults and data visibility helps enable effective analytics without
having to work around the IO stack semantics. While these and other
contributions are valuable, similar efforts made elsewhere may offer attractive
alternatives or differing semantics that could yield a more feature rich
environment with little to no additional overhead. For example, the Doubly
Distributed Transactions (\DDTns) protocol offers an alternative approach for
incorporating transactional semantics into the data path. Another project,
PreDatA, examined how to get the best throughput for data operators and may
offer additional insights into further refinements of the Burst Buffer concept.

This paper examines some of the choices made by the Fast Forward team and
compares them with other options and offers observations and suggestions based
on these other efforts.  This will include some non-core contributions of other
projects, such as some of the demonstration metadata and data storage
components generated while implementing \DDTns, to make suggestions that may
help the next generation design for how the IO stack works as a whole.

@@

The overall design seeks to offer high availability, byte-granular,
multi-version concurrency control. Through the use of a copy-on-write style
mechanism, multiple versions of an object can be stored in potentially greatly
reduced space. It assumes the client interface will be through an IO library
affording a more complicated interface that offers richer functionality
requiring only minimal end-user code changes.  Managing most data access in a
platform-local layer rather than requiring writing to centralized storage will
better support the performance and energy requirements of extreme scale
integrated application workflows. At a more detailed view, the various layers
of the IO stack each contribute different functionality and performance
implications.

The basic architecture incorporates five layers. The top layer is a high level
IO library, such as the demonstration HDF-5 library~\cite{hdf5}. The intent is
to only have access to the storage stack through such an API to manage the
complexity of working with the lower layers. Below the user API is an IO
forwarding layer that redirects IO calls from the compute nodes to the IO
dispatching layer.  This IO forwarding layer is analogous to the function of
the IO nodes in a BlueGene machine. The next two layers have considerable
functionality. The IO dispatcher (IOD) serves as the primary storage interface
for the IO stack and is the only way to reach the persistent storage array in
lower layers. The Distributed Application Object Storage (DAOS) layer serves as
the persistent storage interface and is intended to be the foundation on which
everything else is built with no dependence on any technologies specified above
it. For example, the IOD layer and burst buffers are not required for DAOS to
operate properly. At the bottom is the Versioning Object Storage Device (VOSD).
It serves as the interface for storing objects of all types efficiently.

@@

The idea of burst buffers were initially explored in the context of data
staging~\cite{abbasi:2007:datatap,Abbasi:2009:datatap,nisar:2008:staging,zheng:2010:predata}.
These initial designs all use extra compute nodes to represent the data storage
buffer given the lack of any dedicated hardware support for this functionality.
The desired outcome of these initial studies is to motivate how such
functionality might be incorporated and the potential benefits.  Later, these
concepts were proposed to be incorporated as part of the IO
stack~\cite{nowoczynski:2008:zest,bent:2012:challenges,bent:2012:burst-buffer}.
The current Fast Forward IOD design recommends incorporating SSDs, but
specifically lists these devices as optional. Unfortunately, not incorporating
burst buffers and the use of SSDs in the IOD layer may be problematic.  First,
the IOD design currently is written assuming burst buffers. This means that the
bulk of IO operations will only hit the IOD layer and proposed functionality,
such as the function shipping, do not discuss the impact on the DAOS layer or
the operators themselves should a burst buffer not be available.  Consider the
important functionality of data rearrangement and doing things like changing
the fast array dimension on shared, spinning media.  The design of DAOS assumes
that it will only be involved when persisting a completed transaction and only
for a fraction of the total transactions created.  Transaction frequency at the
IOD layer can be higher since IOD does not ``flatten'' transactions. Flattening
is the process of transferring the set of copy-on-write changes from
transaction/epoch $n$ to transaction $n+m$ into the DAOS layer. The DAOS layer
mirrors the copy-on-write functionality, but stores the differences between
epochs rather than individual transactions. The primary difference is that at
the DAOS layer, data is stored in large, logically contiguous chunks to manage
the metadata load and hopefully improve read performance.

@@

By concentrating the Burst Buffers and function shipping into the storage
stack, three problems arise.  First, the amount of network bandwidth, IO
bandwidth, and compute power consumed for example operations from a single
application is likely to completely monopolize the IOD processes. Second, if
space and time partitioning is used instead, the functionality risks being too
small to be useful. Third, the long-term hardware performance advantage for
SSDs is questionable. Recent studies have shown that the erase-before-write
and interference between reading and writing with flash-based SSDs can cause
severe performance problems~\cite{skourtis:2013:ssd-performance}. The inclusion
of an optional use SSD layer in the new Trinity machine at Los Alamos will
offer a test bed to determine how likely these observed problems would affect a
production extreme scale platform.

@@

Another concern that is acknowledged, but no thought has been applied to, is
the requirement that a single IOD process of the set assigned to an application
is the master for any operation.  Should the number of concurrent applications
exceed the available nodes, sharing an IOD process will be required. The
requirements both in terms of scheduling and resource management were cut from
the project due to funding limitations. Since partitioning of the IOD processes
for exclusive use by particular applications is the assumed operating mode,
should insufficient IOD resources be available, either a job could be delayed
or IOD resources could be reallocated from a different process could be
redeployed for use by the new job. Handling resilience concerns for the IOD
processes must also be address.  These sorts of considerations still need to be
made for a full production system.

@@

\subsubsection{Design Philosophy}
Undocumented, but inherent in the design of these transactions is how faults
are detected. The initial design assumes the current Lustre fault detection
mechanism that can determine if a process or node is no longer reachable. This
detection happens at the DAOS layer and when a fault is detected, the rollback
process is pushed up to the IOD layer for all non-persisted or non-committed
transactions. This defines how a fault will be detected and what will trigger a
passive fault recovery (i.e., transaction abort).

@@

\subsection{Metadata Management}
Metadata management has been a perennial challenge for parallel storage
systems.  Eliminating metadata management as a special case and instead
treating it just as data is a central design goal of the Fast Forward project.
This is a hybrid approach to metadata management that is half-way between
providing no inherent metadata support and having a fully integrated, but
separate metadata management system.
